---
title: "The Effect of Vaccines on COVID-19"
subtitle: "Lily Eichler and Feng Wan"
format: 
  revealjs: 
    embed-resources: true
    theme: "night" 
editor: 
  markdown: 
    wrap: 72
---

## Introduction to the Topic and Our Motivation

COVID-19 has been an incredibly prominent aspect of all of our daily
lives for the past three years. It has resulted in millions of deaths
and innumerable cultural and societal changes. The development and
distribution of billions of vaccines in response have resulted in a
return to a reality that somewhat resembles pre-pandemic life. **Our
study aimed to determine whether vaccines have had a measurable effect
on the number of new COVID-19 cases each day, what this effect is, and
how it differs in different regions around the world.**

## Introduction to Our Data

Our data is a combination of the COVID-19 case data from the World
Health Organization, the vaccine distribution data from the Our World in
Data Organization, and data on the world population by country in 2020
from kaggle.com. We combined the three data sets to examine the effect
of new vaccines distributed on new COVID-19 cases per 1000 people of the
population of each region of the world in 2021 and 2022.

## Hypotheses

Our null hypothesis is that there is no effect of the distribution of
vaccines on new confirmed COVID-19 cases for any region of the world.
H~0~: β = 0.

Our alternative hypothesis is that there is an effect of the
distribution of vaccines on new confirmed COVID-19 cases for at least
one region of the world. H~A~: β != 0.

Our significance level will be equal to 0.05.

## COVID-19 Cases vs. Vaccines

```{r}
#| message: false
#| warning: false
#| echo: false
# Loading the necessary packages and data sets

library(tidyverse)
library(readr)
library(tidymodels)
library(lubridate)
library(knitr)

case <- read_csv("Data/covid_cases.csv")
vaccines <- read_csv("Data/owid-covid-data.csv")
population <- read_csv("Data/population_by_country_2020.csv")
```

```{r}
#| message: false
#| warning: false
#| echo: false
# Mutating, tidying, and joining the data sets as necessary for our analysis

case <- case %>%
  mutate(year = year(Date_reported)) %>%
  mutate(month = month(Date_reported)) %>%
  mutate(day = day(Date_reported)) %>%
  mutate(region = `WHO_region`) %>%
  filter(region != "NA") %>%
  mutate(region = factor(region)) %>%
  mutate(region = fct_relevel(region, "Other", after = 0))

vaccines <- vaccines %>%
  filter(!is.na(total_vaccinations_per_hundred)) %>%
  mutate(year = year(date)) %>%
  mutate(month = month(date)) %>%
  mutate(day = day(date)) %>%
  select(location, date, year, month, day, total_vaccinations_per_hundred)

population <- population %>%
  mutate(Country = `Country (or dependency)`) %>%
  mutate(population = `Population (2020)`) %>%
  select(Country, population)

case <- case %>%
  left_join(population)

full_data <- case %>%
  left_join(vaccines, by = c("Country" = "location", "year", "month", "day")) %>%
  filter(!is.na(total_vaccinations_per_hundred)) %>%
  group_by(Country) %>%
  mutate(lag_vaccine = lag(total_vaccinations_per_hundred, n = 14)) %>%
  filter(!is.na(lag_vaccine)) %>%
  group_by(Country, year, month) %>%
  slice(n = n())

full_data <- full_data %>%
  group_by(Country) %>%
  mutate(new_cases = Cumulative_cases - lag(Cumulative_cases, n = 1), 
         new_vaccine = total_vaccinations_per_hundred - lag(total_vaccinations_per_hundred, n = 1)) %>%
  filter(!is.na(new_cases), !is.na(new_vaccine)) %>%
  mutate(new_case_per_10000 = new_cases * 10000 / population) %>%
  select(Date_reported, Country, year, month, day, region, new_case_per_10000, new_vaccine, population) %>%
  mutate(log_new_case = log(new_case_per_10000 + 0.00001))
```

```{r}
#| message: false
#| warning: false
#| echo: false
# Plotting the relationship between vaccines and cases with a logarithm regression line

full_data %>%
  ggplot(aes(x = new_vaccine, y = log_new_case)) +
  geom_jitter(alpha = 0.5) +
  geom_smooth(se = FALSE, method = "lm") +
  labs(title = "Logged COVID-19 Cases Per Capita vs. COVID-19 Vaccinations Per Hundred",
       subtitle = "By Region (2021-2022)",
       x = "New COVID-19 Vaccinations", 
       y = "New COVID-19 Cases Per 10000 People") + theme_minimal() + facet_wrap(~ region)
```

## Residual Plots

::: columns
::: {.column width="50%"}
```{r}
#| message: false
#| warning: false
#| echo: false
# Plotting the spread of the residuals around the linear regression line

linear_reg <- linear_reg() %>%
  set_engine("lm") %>%
  fit(new_case_per_10000 ~ new_vaccine + day + month + year + region, data = full_data)

linear_reg_aug <- augment(linear_reg$fit)

ggplot(linear_reg_aug, mapping = aes(x = .fitted, y = .resid)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0, color = "gray", lty = "dashed") +
  labs(title = "Linear Residual Plot", x = "Predicted New Cases", y = "Residuals") + theme_minimal()
```

::: incremental
-   Shows a linear pattern
-   Not ideal because it is not randomly scattered
:::
:::

::: {.column width="50%"}
```{r}
log_reg <- linear_reg() %>%
  set_engine("lm") %>%
  fit(log(new_case_per_10000 + 0.00001) ~ new_vaccine + day + month + year + region, data = full_data)

log_reg_aug <- augment(log_reg$fit)

ggplot(log_reg_aug, mapping = aes(x = .fitted, y = .std.resid)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0, color = "gray", lty = "dashed") +
  labs(x = "Predicted New Cases", y = "Residuals",
       title = "Logarithm Residual Plot") + theme_minimal()
```

::: incremental
-   Randomly scattered around the regression line
-   Relatively ideal in comparison with the linear residual plot
:::
:::
:::

## Null Distribution {.smaller}

```{r}
#| warning: false
#| out-width: 100%
# Plotting the logarithm regression line

log_reg <- linear_reg() %>%
  set_engine("lm") %>%
  fit(log_new_case ~ new_vaccine + day + month + year + region, data = full_data)

log_reg_tidy <- log_reg %>%
  tidy()
```

```{r}
#| warning: false
#| out-width: 100%
# Calculating the observed fit, simulating the null distribution for vaccines, and calculating the p-value

obs_fit <- full_data %>%
  specify(log_new_case ~ new_vaccine + day + month + year + region) %>%
  fit()

set.seed(144)
null_fits <- full_data %>%
  specify(log_new_case ~ new_vaccine + day + month + year + region) %>%
  hypothesize(null = "independence") %>%
  generate(reps = 1000, type = "permute") %>%
  fit()

filtered_null_fits <- null_fits %>%
  filter(term %in% c("new_vaccine"))
attr(filtered_null_fits, "fitted") <- TRUE
attr(filtered_null_fits, "type") <- "permute"
  
filtered_null_fits %>%
visualize() + shade_p_value(obs_fit, direction = "both")

p_value <- get_p_value(null_fits, obs_stat = obs_fit, direction = "both")
```

$\hat {NewCase} = `r log_reg_tidy[1, 2]` + \bar {NewVaccine} \times `r exp(log_reg_tidy[2, 2])` + .....$

::: incremental
-   From the null distribution, we calculated a p-value of 0.068
-   Since the p-value is greater than the significance level of 0.05, we
    fail to reject the null hypothesis
-   We cannot claim that there is an effect of COVID-19 vaccines on
    cases for any region of the world
:::

## Confidence Interval

::: incremental
-   The 95% confidence interval is (0.00296, 0.0256)
-   We are 95% confident that the true change of number of cases caused
    by the number of vaccines lies between the values of 0.00296 and
    0.0256
-   This doesn't support the null distribution test, as we fail to
    reject the null hypothesis, suggesting that the true change of cases
    due to vaccines is equal to 0
-   The confidence interval does not include 0 (though it does include
    some values very close to it), so the two tests do not agree
:::

## Conclusions and Future Work

::: incremental
-   In the future, we will:
-   Account for more confounding variables
-   Change the lagged function based on different potency periods of
    different types of vaccines
-   Set a more specific alternative hypothesis (along with a more
    specific null hypothesis to match)
:::
